---
title: "Class exercise week 10 - Designing weights"
author: "Peter Lugtig"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survey)
library(foreign)
```

## Question 1: diagnostics
Consider that you succeeded to draw a relatively large and perfectly random sample from your target population. This means that with fully observed data 
(the 1517 cases in the file “unit non response information.RDS”), the sample would be representative for the population (no worries about coverage and sampling errors for now). 
However, not every respondent that was addressed was willing to fill in your questionnaire, so unfortunately there is ‘unit non-response’. This possibly creates bias. 
Luckily, we still have some basic information about the non-responders, so it is possible to weight back to the full sample (representing the target population).

The first part of the exercise is about checking for the potential bias (a). In the second part (b), you are asked to answer some research-questions correcting for this possible bias.
a.	Open the data file “unit non response information.RDS”. In this file you find all available information about both responders and non-responders. 
Examine if there are any large differences between respondents and non-respondents with regard to region, gender, education, race, and age. In other words: are there strong predictor of response. The following variables are available:

id - Respondent id  
region - 1= Northeast USA, 2= Southern US, 3= West 
sex - 1 = male, 2= female
educ - higher number mean higher levels
race - 1 = white, 2 = black, 3 = other (hispanic)
age - Age in years
response - 0 = nonresponse, 1 = response

```{r check data, include=FALSE}
data <- readRDS("unit non response information.RDS") # read data in as data frame

# check for differences in respondenrs/NR 
boxplot(data$age~data$response) # small diff in age
t.test(data$age~data$response) # small diff in age
chisq.test(table(data$sex,data$response)) # large diff in sex
chisq.test(table(data$region,data$response)) # small diff in region
chisq.test(table(data$race,data$response)) # small diff in race
chisq.test(table(data$educ,data$response)) # small diff in region
# error in education variable
summary(data$educ) # missing value codes wrongly coded
boxplot(data$educ~data$response) # no diff in education

# conclusion. Large diff in sex (gender), smaller diffs for region,race and age. 
prop.table(table(data$sex,data$response)) #only this diff seems highly relevant.
```

## Question 2: Creating a weight    
Two variables were selected for the subsequent weighting exercises: sex and race. Consider the sample to be unbiased with respect to all other variables. 
Compute the Nonresponse weight for the variable sex and race. Compute two weights (for each variable one), and then multiply both weights to achieve one nonresponse weights in your dataset. The easiest way to do this is to run a glm() with a logit link function, and then add the fitted.values (these are the propensity scores) to your dataset. The inverse of these propensity scores is your weight.

```{r create weights, include=FALSE}
# what type of weighting to use?
# different possibilities, here propensity score weights
# want to do linear weights (see below)? 
# create separate weight for each variable, and then multiply them
results <- glm(data$response~data$sex+data$region+data$race+data$age, family=binomial(),
                  data=data)
summary(results)
data <- cbind(fitted.values(results),data) # attach it to original dataset
#boxplot(data$fitted.values) # how are prop-scores distributed?
data$weights <- (1/data$fitted.values)
boxplot(data$weights)
# Note that the average nonresponse weight is now about 1.8, which is equal to the inverse of the response rate (1517/832). 
#That is, the weights now indicate for how many respondents each case stands

```

We now want to add the weights not to the dataset that contains respondents and nonrespondents, but only to the dataset with respondents. The code below loads the datafile "responders.RDS", and then add the weights you created in the previous question to the respondent dataset. 
```{r merge datasets}
# now merge with substantive data 
respdata <- readRDS("responders data.RDS") # read data in as data frame
finaldata <-merge(respdata,data)
```  

## Question 3: explore other ways to create weights. 
There are otherways to weights. Last week we discussed the following four methods

1.	Using poststratification (use the lecture slides about nonresponse). The basic idea is here that you compute a weight for every cell in the multi-way crosstable of your weight-variables.  
2. Using linear weighting using population totals
3. Using a propensity based model using population distribution
4. Using  Raking

The general syntax for these methods look like this:
Poststratification:  
- postStratify(design= svy.unweighted, strata = sexrace, population=sexrace.dist)  

Linear qeighting:  
-	calibrate(design=svy.unweighted,formula=~sex+race, population=c(sex.dist,race.dist), calfun=c(“linear”))  

Propensity score weighting (using population totals)
-	calibrate(design=svy.unweighted,formula=~sex+race, population=c(sex.dist,race.dist), calfun=c(“logit”))  

Raking:  
- rake(design=svy.unweighted,sample.margins=list(~sex,~race), population.margins=list(sex.dist,race.dist))  

Explore each of the weighting methods. Adjust the code below so you also include the variable "region". You may also add other variables, just to try out how these methods work.
```{r do weights, message=F, results='hide'}
# create sexrace crosstable
svy.unweighted <-svydesign(ids=~id, data=respdata)

sexrace<- as.data.frame(table(respdata$sex,respdata$race))
sexrace.dist <-xtabs(~sex+race, data=data)

poststratdesign <- postStratify(design= svy.unweighted, strata =~sex+race, population=sexrace.dist) 

# for Calibration (linear weighting)
sex.dist <- as.data.frame(table(data$sex))
colnames(sex.dist) <- c("sex", "Freq")
race.dist <- as.data.frame(table(data$race))
colnames(race.dist) <- c("race", "Freq")
table(data$sex,data$race)
pop.totals<-c('(Intercept)'=545+719,sexFemale=719+133+29,raceBlack=133+71,raceOther=20+29) 
# you need to get the totals as numbers.

linearweightdesign <- calibrate(design=svy.unweighted,formula=~sex+race, 
                                population=pop.totals, calfun=c("linear")) #linear weighting
propweightdesign  <- calibrate(design=svy.unweighted,formula=~sex+race, 
                               population=pop.totals, calfun=c("logit"), bounds=c(0.3,3)) 
#inverse probability (propensity) 
# a note about the bounds here -> they are needed to make sure the model converges. 
# I trmmed at 3, but you can choose other values 
# remember bias/variance trade-off?

rakeddesign  <- rake(design=svy.unweighted,sample.margins=list(~sex,~race),
            population.margins=list(sex.dist,race.dist)) 
```

## Question 4: does it matter?.
Compute the mean and variance in ‘income’ for the respondents without weights and then using the (three) weights that you created in the previous question. Do you find any differences between the weighted and unweighted estimates? And does the weighting method make a difference? Do you think that variance inflation due to weighting is a problem here?

```{r analysis weighted and unweighted, include=F}
mean(finaldata$income, na.rm=T) #2183
# 2183.9
# weighted
svymean(~income, design=poststratdesign, na.rm=T) # mean = 2198.6, SE=20.45
svymean(~income, design=linearweightdesign, na.rm=T) # mean = 2177, se=19.13
svymean(~income, design=propweightdesign, na.rm=T) # mean=2177, se=19.137
svymean(~income, design=rakeddesign, na.rm=T) #mean=21999, se=20.44
# results will slightly differ if you used different variables.
```

## Question 5 (optional)
How large is the difference in income between men and women under all weightingscenarios?

```{r analysis, results='hide'}
t.test(finaldata$income~ finaldata$sex)
svyttest(income~sex, design=poststratdesign,family=gaussian)
# etc.
```

## Question 6	(optional):
To get a more real-life situation, I looked up some employment statistics in the USA. The variable ‘employm’ indicates for respondents whether they have a job (1) or not (0). The population distribution for employment is as follows:

Employed: 48,2%
Not employed: 51.8%

Use raking to correct for nonresponse using this additional variable . Use employm + the variables you used earlier in this exercise. Inspect the effects of using the additional variables on the mean and variance of ‘income’.

	Example code:
rake(design = svy.unweighted,(sample.margins = list(~sex, ~race, ~employm),(population.margins = list(sex.dist, race.dist, employ.dist))

```{r raking with employment, include=F}
employ.dist <- as.data.frame(as.factor(c('employed','unemployed'))) # factors as in dataset
employ.dist[,2] <- as.numeric(.482,.518) # add a column with the numbers
colnames(employ.dist) <- c("employm"," Freq")
rakeddesign2 <- rake(design = svy.unweighted,sample.margins = list(~sex, ~race, ~employm),population.margins = list(sex.dist, race.dist, employ.dist)) # rake
svymean(~income, design=rakeddesign2, na.rm=T) #mean=2200.6, se=20.9
```


-- End of document --